---
title: "3D Correlation Space"
format: html
editor: visual
execute:
  echo: false
  warning: false
---

```{r}
#| echo: false
#devtools::install_github("rmcelreath/rethinking@slim")
pacman::p_load(data.table, dplyr, plotly, rethinking)
```

# Goal

-   **Ultimate goal**: Develop an approach that allows us to claim "With 95% confidence, this confidence region contains the true set".

-   **Temporary goal**: In reality, the true set is unobserved; we only observe data. Since many causal discovery algorithms operates on the data correlation matrix $R_{obs}$, our intermediate goal is to generate perturbations around $R_{obs}$, aiming to represent the sampling distribution of the correlation estimator under repeated sampling. We then use this approximation to construct a confidence region for the underlying population correlation structure, and study how causal discovery perform across this region.

    Some questions to think and ask:

    1.  What is our scope, what assumptions are we making:

        We assume infinite sample at this stage, distribution of pearson correlation estimators is asymptotically normal. Want a frequentist approach.

    2.  How to generate perturbations around $R_{obs}$ ?

    3.  How do measure the distance between each perturbed sample $R^{\prime}_{obs}$ with $R_{obs}$ ?

    4.  

# Visualizing Correlation Space

Approaches:

1.  Construct x, y, z $\in$ \[-1,1\]. Check which combination is positive semi-definite using determinants. Plot results

2.  Leverage LKJ distribution, sample N samples and plot.

    The LKJ distribution provides

```{r}
#| echo: false
x<- y<- z<- seq(-1,1, length= 30)   #r12, r13, r23
dt<- expand.grid(x=x,y=y,z=z) %>% as.data.table
dt[,psd:= 1-x^2 -y^2-z^2- 2*x*y*z]
dt<- dt[psd>0,]
p0<- plot_ly(data= dt, x = ~x, y = ~y, z = ~z, type = "scatter3d",
  mode = "markers",
  marker = list(size = 2)) %>% layout(
  title = list(
    text = "Approach 1: Valid 3-d correlation matrices by PSD",
    x = 0.5,       # 0 = left, 0.5 = center, 1 = right
    xanchor = "center",
    yanchor = "top",
    font = list(size = 20, color = "black")
  ),
  scene = list(
    xaxis = list(title = "r12"),
    yaxis = list(title = "r13"),
    zaxis = list(title = "r23")
  )
)
p0
```

```{r}
#| echo: false
N <- 3000
samp1 <- rlkjcorr(N, K = 3, eta = 0.1) #eta parameter represents density of sampling
samp2<- rlkjcorr(N, K = 3, eta = 1)
samp3 <- rlkjcorr(N, K = 3, eta = 10)

df1 <- data.frame(x=samp1[, 2, 1], y=samp1[, 3, 1], z=samp1[, 3, 2])
df2 <- data.frame(x=samp2[, 2, 1], y=samp2[, 3, 1], z=samp2[, 3, 2])
df3 <- data.frame(x=samp3[, 2, 1], y=samp3[, 3, 1], z=samp3[, 3, 2])

p1<- plot_ly(df1,  x = ~x, y = ~y, z = ~z, type = "scatter3d", 
  mode = "markers",   marker = list(size = 3), name="eta=0.1 (sample towards edges") %>%
  layout(title = "Approach 2: LKJ-distribution")
p2<- plot_ly(df2,  x = ~x, y = ~y, z = ~z, type = "scatter3d",color = I("lightblue"),
  mode = "markers",   marker = list(size = 3), name="eta=1 (uniform distribution)") 
p3<- plot_ly(df3,  x = ~x, y = ~y, z = ~z, type = "scatter3d",
  mode = "markers",   marker = list(size = 3), name="eta=10 (sample towards I") 
subplot(p1, p2, p3, nrows = 1, margin=0.05)

```

## $\varepsilon$-ball around $R_{pop}$

```{r}
# Generate N samples around Rpop with a given noise-level
r.eps<- function(Rpop, N=1, noise= 0.1, pd=FALSE){
  p= ncol(Rpop)
  lapply(1:N, function(i){
    rE<- rnorm(p*(p-1)/2,0,1) # sampling eps ball with radius 1
    rE<- rE/ sqrt(sum(rE^2)) #normalization
    rE<- rE* noise
    E<- diag(0,p,p)
    
    E[upper.tri(E)]<- rE
    E <- (E + t(E))
    Ri<- Rpop+E
    
    if(pd==TRUE) Ri<- Matrix::nearPD(Ri, corr=TRUE)$mat
    return(Ri)
  })
}


# Flatten correlation matrix for plotting purpose
R_to_coords <- function(R) {
  data.frame(
    r12 = R[1,2],
    r13 = R[1,3],
    r23 = R[2,3]
  )
}

```

```{r plotRi}
Rpop<- rlkjcorr(1, K = 3, eta = 30)
R1<- r.eps(Rpop, N=200, noise= 0.1)
R2<- r.eps(Rpop, N=200, noise= 0.3)
R3<- r.eps(Rpop, N=200, noise= 0.5)
R3a<- r.eps(Rpop, N=200, noise= 0.5, pd = TRUE)

df_Rpop <- R_to_coords(Rpop)
df_R1 <- do.call(rbind, lapply(R1, R_to_coords))
df_R2 <- do.call(rbind, lapply(R2, R_to_coords))
df_R3 <- do.call(rbind, lapply(R3, R_to_coords))
df_R3a <- do.call(rbind, lapply(R3a, R_to_coords))

p20 <- p2 %>%
  # Add Rpop as a red marker
  add_trace(
    data = df_Rpop,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 6, color = "red"),
    name = "Rpop"
  ) %>%
  # Add Ri matrices as blue markers
  add_trace(
    data = df_R1,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "orange"),
    name = "noise=0.1"
  ) %>%
  add_trace(
    data = df_R2,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "green"),
    name = "noise=0.3"
  ) %>%
  add_trace(
    data = df_R3,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "pink"),
    name = "noise=0.5"
  ) %>%
  add_trace(
    data = df_R3a,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "purple"),
    name = "noise=0.5 + nearest pos-definite"
  ) %>%
  layout(
    legend = list(itemsizing = "constant"),
    title = list(
    text = "When Rpop is close to I",
    x = 0.5,       # 0 = left, 0.5 = center, 1 = right
    xanchor = "center",
    yanchor = "top",
    font = list(size = 20, color = "black")
  
  )
  )
p20
```

```{r}
N=500
Rpop<- rlkjcorr(1, K = 3, eta = 0.1)
R1<- r.eps(Rpop, N, noise= 0.1)
R2<- r.eps(Rpop, N, noise= 0.3)
R3<- r.eps(Rpop, N, noise= 0.5)

df_Rpop <- R_to_coords(Rpop)
df_R1 <- do.call(rbind, lapply(R1, R_to_coords))
df_R2 <- do.call(rbind, lapply(R2, R_to_coords))
df_R3 <- do.call(rbind, lapply(R3, R_to_coords))

p20 <- p2 %>%
  # Add Rpop as a red marker
  add_trace(
    data = df_Rpop,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 6, color = "red"),
    name = "Rpop"
  ) %>%
  # Add Ri matrices as blue markers
  add_trace(
    data = df_R1,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "orange"),
    name = "noise=0.1"
  ) %>%
  add_trace(
    data = df_R2,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "green"),
    name = "noise=0.3"
  ) %>%
  add_trace(
    data = df_R3,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "pink"),
    name = "noise=0.5"
  ) %>%
  layout(
    legend = list(itemsizing = "constant"),
    title = list(
    text = "When Rpop is close to boundary",
    x = 0.5,       # 0 = left, 0.5 = center, 1 = right
    xanchor = "center",
    yanchor = "top",
    font = list(size = 20, color = "black")
  
  )
  )
p20
```

# Asymptotic Distribution of R

# 
