---
title: "3D Correlation Space"
format: html
editor: visual
execute:
  echo: false
  warning: false
---

```{r}
#| echo: false
#devtools::install_github("rmcelreath/rethinking@slim")
pacman::p_load(data.table, dplyr, plotly, rethinking)
```

# Goal

-   **Ultimate goal**: Develop an approach that allows us to claim "With 95% confidence, this confidence region contains the true set".

-   **Temporary goal**: In reality, the true set is unobserved; we only observe data. Since many causal discovery algorithms operates on the data correlation matrix $R_{obs}$, our intermediate goal is to generate perturbations around $R_{obs}$, aiming to represent the sampling distribution of the correlation estimator under repeated sampling. We then use this approximation to construct a confidence region for the underlying population correlation structure, and study how causal discovery perform across this region.

-   Some questions:

    1.  What is our scope, what assumptions are we making:

        We assume infinite sample at this stage, distribution of pearson correlation estimators is asymptotically normal. Want a frequentist approach.

    2.  What is the correlation space?

        $p$- dimensional elliptope

    3.  How to generate perturbations around $R_{obs}$ ?\
        Using asymptotic theory, we get the asymptotic distribution of $R_{obs}$ and generate samples from the null distribution.

    4.  How to measure the distance between each perturbed sample $R^{\prime}_{obs}$ with $R_{obs}$ ?

        Mahalanobis distance

    5.  

# Notations

$R$: True R, unknown

$R_{obs}$: Observed R, typically estimate from sample data

$R^{\prime}_{obs}$: Use superscript "$\prime$ " to denote a perturbed object

$\rho$: $vech(R)$, the unique off-diagonal elements of $R$ matrix.

$\hat\rho$: $vech(R_{obs})$, the unique off-diagonal elements of $R_{obs}$

$\hat\rho^{\prime}$: $vech(R^{\prime}_{obs})$, the unique off-diagonal elements of $R^{\prime}_{obs}$

$\mu$: population mean

$\Sigma$: Covariance matrix

$\Psi$: Covariance matrix of pearson correlation coefficient

$d$: number of variables in data

$p$: number of parameters (length of $\rho$) = $\frac{d(d-1)}{2}$

# Visualizing Correlation Space

## Correlation space for 3 variables (d=3)

Sample correlation matrices from LKJ distribution, which uses $\eta$ to parametrize *centrality*: larger $\eta$, closer to the identity matrix.

```{r}
#| echo: false
N <- 3000
samp1 <- rlkjcorr(N, K = 3, eta = 0.1) #eta parameter represents density of sampling
samp2 <- rlkjcorr(N, K = 3, eta = 1)
samp3 <- rlkjcorr(N, K = 3, eta = 10)

df1 <- data.frame(x=samp1[, 2, 1], y=samp1[, 3, 1], z=samp1[, 3, 2])
df2 <- data.frame(x=samp2[, 2, 1], y=samp2[, 3, 1], z=samp2[, 3, 2])
df3 <- data.frame(x=samp3[, 2, 1], y=samp3[, 3, 1], z=samp3[, 3, 2])

p1<- plot_ly(df1,  x = ~x, y = ~y, z = ~z, type = "scatter3d", 
  mode = "markers",   marker = list(size = 3), name="eta=0.1 (towards boundary") %>%
  layout(title = "Correlation Space")
p2<- plot_ly(df2,  x = ~x, y = ~y, z = ~z, type = "scatter3d",color = I("lightblue"),
  mode = "markers",   marker = list(size = 3), name="eta=1 (uniform)") 
p3<- plot_ly(df3,  x = ~x, y = ~y, z = ~z, type = "scatter3d",
  mode = "markers",   marker = list(size = 3), name="eta=10 (towards Identity") 
subplot(p1, p2, p3, nrows = 1, margin=0.05)

```

## $\varepsilon$-ball around $R_{pop}$

```{r}
# Generate N samples around Rpop with a given noise-level
r.eps<- function(Rpop, N=1, noise= 0.1, pd=FALSE){
  p= ncol(Rpop)
  lapply(1:N, function(i){
    rE<- rnorm(p*(p-1)/2,0,1) # sampling eps ball with radius 1
    rE<- rE/ sqrt(sum(rE^2)) #normalization
    rE<- rE* noise
    E<- diag(0,p,p)
    
    E[upper.tri(E)]<- rE
    E <- (E + t(E))
    Ri<- Rpop+E
    
    if(pd==TRUE) Ri<- Matrix::nearPD(Ri, corr=TRUE)$mat
    return(Ri)
  })
}


# Flatten correlation matrix for plotting purpose
R_to_coords <- function(R) {
  data.frame(
    r12 = R[1,2],
    r13 = R[1,3],
    r23 = R[2,3]
  )
}

```

```{r plotRi}
Rpop<- rlkjcorr(1, K = 3, eta = 30)
R1<- r.eps(Rpop, N=200, noise= 0.1)
R2<- r.eps(Rpop, N=200, noise= 0.3)
R3<- r.eps(Rpop, N=200, noise= 0.5)
R3a<- r.eps(Rpop, N=200, noise= 0.5, pd = TRUE)

df_Rpop <- R_to_coords(Rpop)
df_R1 <- do.call(rbind, lapply(R1, R_to_coords))
df_R2 <- do.call(rbind, lapply(R2, R_to_coords))
df_R3 <- do.call(rbind, lapply(R3, R_to_coords))
df_R3a <- do.call(rbind, lapply(R3a, R_to_coords))

p20 <- p2 %>%
  # Add Rpop as a red marker
  add_trace(
    data = df_Rpop,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 6, color = "red"),
    name = "Rpop"
  ) %>%
  # Add Ri matrices as blue markers
  add_trace(
    data = df_R1,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "orange"),
    name = "noise=0.1"
  ) %>%
  add_trace(
    data = df_R2,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "green"),
    name = "noise=0.3"
  ) %>%
  add_trace(
    data = df_R3,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "pink"),
    name = "noise=0.5"
  ) %>%
  add_trace(
    data = df_R3a,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "purple"),
    name = "noise=0.5 + nearest pos-definite"
  ) %>%
  layout(
    legend = list(itemsizing = "constant"),
    title = list(
    text = "When Rpop is close to I",
    x = 0.5,       # 0 = left, 0.5 = center, 1 = right
    xanchor = "center",
    yanchor = "top",
    font = list(size = 20, color = "black")
  
  )
  )
p20
```

```{r}
N=500
Rpop<- rlkjcorr(1, K = 3, eta = 0.1)
R1<- r.eps(Rpop, N, noise= 0.1)
R2<- r.eps(Rpop, N, noise= 0.3)
R3<- r.eps(Rpop, N, noise= 0.5)

df_Rpop <- R_to_coords(Rpop)
df_R1 <- do.call(rbind, lapply(R1, R_to_coords))
df_R2 <- do.call(rbind, lapply(R2, R_to_coords))
df_R3 <- do.call(rbind, lapply(R3, R_to_coords))

p20 <- p2 %>%
  # Add Rpop as a red marker
  add_trace(
    data = df_Rpop,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 6, color = "red"),
    name = "Rpop"
  ) %>%
  # Add Ri matrices as blue markers
  add_trace(
    data = df_R1,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "orange"),
    name = "noise=0.1"
  ) %>%
  add_trace(
    data = df_R2,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "green"),
    name = "noise=0.3"
  ) %>%
  add_trace(
    data = df_R3,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "pink"),
    name = "noise=0.5"
  ) %>%
  layout(
    legend = list(itemsizing = "constant"),
    title = list(
    text = "When Rpop is close to boundary",
    x = 0.5,       # 0 = left, 0.5 = center, 1 = right
    xanchor = "center",
    yanchor = "top",
    font = list(size = 20, color = "black")
  
  )
  )
p20
```

# Wald-Type Confidence Regions for the Correlation Vector $\rho$

## Asymptotic Distribution

Under standard regularity conditions: normality and finite asypmtotic variance (finite fourth-order moment $$
  \sqrt{n}(\hat{\rho}-\rho) \xrightarrow{d} N(0,\Psi)
$$where:

-   diagonal of $\Psi$: $\psi_{jk} = 1-\rho_{jk}^2$

-   off-diagonal of $\Psi$:

$$
\psi_{jk,hm}
\;=\;
N\,\sigma_{jk,hm}
\;=\;
\frac{1}{2}\Big[
(\rho_{jh}-\rho_{jk}\rho_{kh})(\rho_{km}-\rho_{kh}\rho_{hm})
\;+\;
(\rho_{jm}-\rho_{jk}\rho_{km})(\rho_{kh}-\rho_{kj}\rho_{jh})
\;+\;
(\rho_{jh}-\rho_{jm}\rho_{mh})(\rho_{km}-\rho_{kj}\rho_{jm})
\;+\;
(\rho_{jm}-\rho_{jk}\rho_{km})(\rho_{kh}-\rho_{km}\rho_{mh})
\Big].
$$

When $\rho$ is unknown, we use plug-in estimate $\hat\rho$

## Confidence Ellipsoid of $\rho$

Let $Z = \Psi^{-\frac{1}{2}}{\sqrt{n}}(\hat{\rho} - \rho)$, we have $Z \xrightarrow{d} N(0,I_p)$. Therefore, $Z^TZ \xrightarrow{d} \chi^2_{p}$ (can be proven using Slutsky Theorem).

The confidence ellipsoid $C_{1-\alpha} = \{ \rho:  \ n(\hat{\rho} - \rho)^T\Psi^{-1}(\hat{\rho} - \rho)\ \leq \chi^{2}_{p, 1-\alpha} \}$, where $\rho$ must satisfy the correlation matrix criteria: PSD.

All perturbed vectors $\hat\rho^{\prime}$ satisfying the inequality are accepted as points inside the ellipsoid.

# Generating random correlation samples around $\rho_0$

What are the approaches:

-   Generate MVN data based on $\hat\rho$ and compute its correlation vector $\hat\rho^{\prime}$
-   Generate $\hat\rho^{\prime}$ from $\hat\rho$ directly:
    -   Sample uniformly from a sphere-cloud

    -   Sample from a wald-cloud

        -   Sample from the full distributional spread around $\rho_0$:

            Let $\rho_0 = vech(R_0)$. For testing $H_0: \rho = \rho_0$, based on asymptotic normality of $\rho$, we have wald-type statistic

            $$
            W = (\hat\rho - \rho_0)^T \Psi(\rho_0)^{-1} (\hat\rho - \rho_0) \sim \chi^2_p
            $$

            This implies we can draw $\hat\rho$ under the null distribution $H_0$:

            $$
            \hat\rho^{(b)} \sim N_p(\rho_0, \frac{\Psi(\rho_0)}{n})
            $$

            Or, equivalently, $\hat\rho^{(b)} = \rho_0 + Lz$, where $z \sim N(0,I_p)$, $L= UD^\frac{1}{2}$ (eigen-decomposition of $\Psi(\rho_0)$.

        -   Sample **uniformly** around $\rho_0$:

            $$
            \hat\rho^{(b)} = \rho_0 + L(\sqrt{c} r^{\frac{1}{p}}u)
            $$

            where $c: \chi^2_{p, 1-\alpha}$, $r$ represent a random radius sample from $Unif(0,1)$, $u$ represents the random unit direction, sample from $N(0,I_p)$ and rescale to have length (norm) 1.

            The role of $c$ is to ensure we sample within some measurable bound (e.g., 95% confidence ellipsoid).

2
