---
title: "3D Correlation Space"
format:
  html:
    css: styles.css
editor: visual
execute:
  echo: false
  warning: false
---

```{r}
#| echo: false
#devtools::install_github("rmcelreath/rethinking@slim")
pacman::p_load(data.table, dplyr, plotly, rethinking)
```

# Goal

-   **Ultimate goal**: Develop an approach that allows us to claim "With 95% confidence, this confidence region contains the true set".

-   **Temporary goal**: In reality, the true set is unobserved; we only observe data. Since many causal discovery algorithms operates on the data correlation matrix $R_{obs}$, our intermediate goal is to generate perturbations around $R_{obs}$, aiming to represent the sampling distribution of the correlation estimator under repeated sampling. We then use this approximation to construct a confidence region for the underlying population correlation structure, and study how causal discovery perform across this region.

-   Some questions:

    1.  What is our scope, what assumptions are we making:

        We assume infinite sample at this stage, distribution of pearson correlation estimators is asymptotically normal. Want a frequentist approach.

    2.  What is the correlation space?

        $p$- dimensional elliptope

    3.  How to generate perturbations around $R_{obs}$ ?\
        Using asymptotic theory, we get the asymptotic distribution of $R_{obs}$ and generate samples from the null distribution.

    4.  How to measure the distance between each perturbed sample $R^{\prime}_{obs}$ with $R_{obs}$ ?

        Mahalanobis distance

    5.  

# Notations

$R$: True R, unknown

$R_{obs}$: Observed R, typically estimate from sample data

$R^{\prime}_{obs}$: Use superscript "$\prime$ " to denote a perturbed object

$\rho$: $vech(R)$, the unique off-diagonal elements of $R$ matrix.

$\hat\rho$: $vech(R_{obs})$, the unique off-diagonal elements of $R_{obs}$

$\hat\rho^{\prime}$: $vech(R^{\prime}_{obs})$, the unique off-diagonal elements of $R^{\prime}_{obs}$

$\mu$: population mean

$\Sigma$: Covariance matrix

$\Psi$: Covariance matrix of pearson correlation coefficient

$d$: number of variables in data

$p$: number of parameters (length of $\rho$) = $\frac{d(d-1)}{2}$

# Visualizing $\rho$ Space

## Correlation space for 3 variables (d=3)

Sample correlation matrices from LKJ distribution, which uses $\eta$ to parametrize *centrality*: larger $\eta$, closer to the identity matrix.

```{r}
#| echo: false
N <- 3000
samp1 <- rlkjcorr(N, K = 3, eta = 0.1) #eta parameter represents density of sampling
samp2 <- rlkjcorr(N, K = 3, eta = 1)
samp3 <- rlkjcorr(N, K = 3, eta = 10)

df1 <- data.frame(x=samp1[, 2, 1], y=samp1[, 3, 1], z=samp1[, 3, 2])
df2 <- data.frame(x=samp2[, 2, 1], y=samp2[, 3, 1], z=samp2[, 3, 2])
df3 <- data.frame(x=samp3[, 2, 1], y=samp3[, 3, 1], z=samp3[, 3, 2])

p1<- plot_ly(df1,  x = ~x, y = ~y, z = ~z, type = "scatter3d", 
  mode = "markers",   marker = list(size = 3), name="eta=0.1 (towards boundary") %>%
  layout(title = "Correlation Space")
p2<- plot_ly(df2,  x = ~x, y = ~y, z = ~z, type = "scatter3d",color = I("lightblue"),
  mode = "markers",   marker = list(size = 3), name="eta=1 (uniform)") 
p3<- plot_ly(df3,  x = ~x, y = ~y, z = ~z, type = "scatter3d",
  mode = "markers",   marker = list(size = 3), name="eta=10 (towards Identity") 
subplot(p1, p2, p3, nrows = 1, margin=0.05)

```

## $\varepsilon$-ball around $R_{pop}$

```{r}
# Generate N samples around Rpop with a given noise-level
r.eps<- function(Rpop, N=1, noise= 0.1, pd=FALSE){
  p= ncol(Rpop)
  lapply(1:N, function(i){
    rE<- rnorm(p*(p-1)/2,0,1) # sampling eps ball with radius 1
    rE<- rE/ sqrt(sum(rE^2)) #normalization
    rE<- rE* noise
    E<- diag(0,p,p)
    
    E[upper.tri(E)]<- rE
    E <- (E + t(E))
    Ri<- Rpop+E
    
    if(pd==TRUE) Ri<- Matrix::nearPD(Ri, corr=TRUE)$mat
    return(Ri)
  })
}


# Flatten correlation matrix for plotting purpose
R_to_coords <- function(R) {
  data.frame(
    r12 = R[1,2],
    r13 = R[1,3],
    r23 = R[2,3]
  )
}

```

```{r plotRi}
Rpop<- rlkjcorr(1, K = 3, eta = 30)
R1<- r.eps(Rpop, N=200, noise= 0.1)
R2<- r.eps(Rpop, N=200, noise= 0.3)
R3<- r.eps(Rpop, N=200, noise= 0.5)
R3a<- r.eps(Rpop, N=200, noise= 0.5, pd = TRUE)

df_Rpop <- R_to_coords(Rpop)
df_R1 <- do.call(rbind, lapply(R1, R_to_coords))
df_R2 <- do.call(rbind, lapply(R2, R_to_coords))
df_R3 <- do.call(rbind, lapply(R3, R_to_coords))
df_R3a <- do.call(rbind, lapply(R3a, R_to_coords))

p20 <- p2 %>%
  # Add Rpop as a red marker
  add_trace(
    data = df_Rpop,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 6, color = "red"),
    name = "Rpop"
  ) %>%
  # Add Ri matrices as blue markers
  add_trace(
    data = df_R1,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "orange"),
    name = "noise=0.1"
  ) %>%
  add_trace(
    data = df_R2,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "green"),
    name = "noise=0.3"
  ) %>%
  add_trace(
    data = df_R3,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "pink"),
    name = "noise=0.5"
  ) %>%
  add_trace(
    data = df_R3a,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "purple"),
    name = "noise=0.5 + nearest pos-definite"
  ) %>%
  layout(
    legend = list(itemsizing = "constant"),
    title = list(
    text = "When Rpop is close to I",
    x = 0.5,       # 0 = left, 0.5 = center, 1 = right
    xanchor = "center",
    yanchor = "top",
    font = list(size = 20, color = "black")
  
  )
  )
p20
```

```{r}
N=500
Rpop<- rlkjcorr(1, K = 3, eta = 0.1)
R1<- r.eps(Rpop, N, noise= 0.1)
R2<- r.eps(Rpop, N, noise= 0.3)
R3<- r.eps(Rpop, N, noise= 0.5)

df_Rpop <- R_to_coords(Rpop)
df_R1 <- do.call(rbind, lapply(R1, R_to_coords))
df_R2 <- do.call(rbind, lapply(R2, R_to_coords))
df_R3 <- do.call(rbind, lapply(R3, R_to_coords))

p20 <- p2 %>%
  # Add Rpop as a red marker
  add_trace(
    data = df_Rpop,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 6, color = "red"),
    name = "Rpop"
  ) %>%
  # Add Ri matrices as blue markers
  add_trace(
    data = df_R1,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "orange"),
    name = "noise=0.1"
  ) %>%
  add_trace(
    data = df_R2,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "green"),
    name = "noise=0.3"
  ) %>%
  add_trace(
    data = df_R3,
    x = ~r12, y = ~r13, z = ~r23,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, color = "pink"),
    name = "noise=0.5"
  ) %>%
  layout(
    legend = list(itemsizing = "constant"),
    title = list(
    text = "When Rpop is close to boundary",
    x = 0.5,       # 0 = left, 0.5 = center, 1 = right
    xanchor = "center",
    yanchor = "top",
    font = list(size = 20, color = "black")
  
  )
  )
p20
```

# Inference for $\rho$

Note: all the below theories have a corresponding z-transformed (variance stabilizing) approach. Details to be added later.

## Asymptotic Distribution

Under standard regularity conditions: normality and finite asypmtotic covariance $\Psi \in R^{p \times p}$ (and also finite fourth-order moment), CLT shows that $$
  \sqrt{n}(\hat{\rho}-\rho) \xrightarrow{d} N(0,\Psi)
$$where:

-   diagonal of $\Psi$: $\psi_{jk} = 1-\rho_{jk}^2$

-   off-diagonal of $\Psi$:

$$
\psi_{jk,hm}
\;=\;
N\,\sigma_{jk,hm}
\;=\;
\frac{1}{2}\Big[
(\rho_{jh}-\rho_{jk}\rho_{kh})(\rho_{km}-\rho_{kh}\rho_{hm})
\;+\;
(\rho_{jm}-\rho_{jk}\rho_{km})(\rho_{kh}-\rho_{kj}\rho_{jh})
\;+\; \\
(\rho_{jh}-\rho_{jm}\rho_{mh})(\rho_{km}-\rho_{kj}\rho_{jm})
\;+\;
(\rho_{jm}-\rho_{jk}\rho_{km})(\rho_{kh}-\rho_{km}\rho_{mh})
\Big].
$$

When $\rho$ is unknown, we use plug-in estimate $\hat\rho$.

<span style="color: red;"> Can we prove that $\hat\Psi \xrightarrow{p} \Psi$ ?</span>

## Confidence region of $\rho$

Let $Z = \Psi^{-\frac{1}{2}}{\sqrt{n}}(\hat{\rho} - \rho)$, we have $Z \xrightarrow{d} N(0,I_p)$. Therefore, by continuous mapping theorem, $Z^TZ \xrightarrow{d} \chi^2_{p}$.

Formally, consider the testing framework:

$H_0: \rho = \rho_0$ vs $H_1: \rho \neq \rho_0$

Under $H_0$,

$$
  \sqrt{n}(\hat{\rho}-\rho_0) \xrightarrow{d} N(0,\Psi)
$$

To test the hypothesis jointly for a $p$-length vector, consider the test statistic

$$
T = n(\hat{\rho} - \rho_0)^T\Psi^{-1}(\hat{\rho} - \rho_0)\ \leq \chi^{2}_{p, 1-\alpha}
$$

We previously show that $T  \xrightarrow{d} \chi^2_{p}$ under $H_0$. Therefore, when $T \geq \chi^2_{p, 1-\alpha}$, we reject $H_0$.

The confidence region can be obtained through test inversion: $C_{1-\alpha} = \{ \rho:  \ n(\hat{\rho} - \rho)^T\Psi^{-1}(\hat{\rho} - \rho)\ \leq \chi^{2}_{p, 1-\alpha} \}$.

All perturbed vectors $\hat\rho^{\prime}$ satisfying the inequality are accepted as points inside the ellipsoid.

When $\Psi$ is unknown, by slutsky theorem, $C_{1-\alpha} = \{ \rho:  \ n(\hat{\rho} - \rho)^T\hat\Psi(\hat\rho)^{-1}(\hat{\rho} - \rho)\ \leq \chi^{2}_{p, 1-\alpha} \}$

Note:

1.  I do think the confidence ellipsoid should be the *intersect* of $C_{1-\alpha}$ <span style="color: red;"> and </span> the PSD criteria.
2.  The plug-in approach is only true when $\hat\Psi \xrightarrow{p} \Psi$ ? , still need to prove this.

### Visualization

::: {style="display:grid; grid-template-columns: 1fr 1fr; gap: 12px;"}
<iframe src="docs/figures/p1.html" width="100%" height="520" style="border:0;"></iframe> <iframe src="docs/figures/p2.html" width="100%" height="520" style="border:0;"></iframe> <iframe src="docs/figures/p3.html" width="100%" height="520" style="border:0;"></iframe> <iframe src="docs/figures/p4.html" width="100%" height="520" style="border:0;"></iframe>
:::

# Generating random samples around $\rho_0$

What are the different approach to generate random samples?

1.  Generate MVN data based on $\hat\rho$ and compute its correlation vector $\hat\rho^{\prime}$

2.  Generate $\hat\rho^{\prime}$ from $\hat\rho$ directly: sample randomly from the full distributional spread around $\rho_0$ (wald-cloud):

    Let $\rho_0 = vech(R_0)$. For testing $H_0: \rho = \rho_0$, based on asymptotic normality of $\hat\rho$, we have wald-type statistic (same as previously mentioned)

    $$
    T = n (\hat\rho - \rho_0)^T \Psi(\rho_0)^{-1} (\hat\rho - \rho_0) \sim \chi^2_p
    $$

    This implies we can draw $\hat\rho$ under the null distribution $H_0$:

    $$
    \hat\rho^{(b)} \sim N_p(\rho_0, \frac{\Psi(\rho_0)}{n})
    $$

    Equivalently, the linear algebra approach:

    $\hat\rho^{(b)} = \rho_0 + Lz$, where $z \sim N(0,I_p)$, $L= \frac{1}{\sqrt{n}}UD^\frac{1}{2}$ (eigen-decomposition of $\Psi(\rho_0)$: $U$ is the eigenvectors , $D$ is the eigenvalue matrix)

    This should give us below statement when $n$ is infinite, $$
    Pr_{\rho_0} (\rho \in C_{1-\alpha}(\hat\rho)) = 1-\alpha 
    $$

Why don't we use the first approach (Generate data then compute correlation) ?

1.  Making extra assumption: This approach assumes the underlying data distribution, which seems to be unnecessary here in our theory. We only need $\hat\rho$ to be normal (asymptotically).
2.  To meet out goal as a less expensive approach for resampling. Sample from $\rho$ directly will be a less expensive approach.

Notes:

1.  Previously discuss a sphere-cloud approach, I modified it to the "wald-cloud"(approach 2) above, this is to consider the covariance of correlation under null.

### Visualization

::: {style="display:grid; grid-template-columns: 1fr 1fr; gap: 12px;"}
<iframe src="plots/p1.html" width="100%" height="520" style="border:0;"></iframe> <iframe src="plots/p2.html" width="100%" height="520" style="border:0;"></iframe> <iframe src="plots/p3.html" width="100%" height="520" style="border:0;"></iframe> <iframe src="plots/p4.html" width="100%" height="520" style="border:0;"></iframe>
:::

# Generating uniform samples inside confidence region of $\rho_0$ 

The goal is to generate **uniform** and **random** samples **within** $C_{1-\alpha}(\rho_0)$ with guarantee: $C_{1-\alpha} = \{ \rho:  \ n(\hat{\rho} - \rho_0)^T\Psi^{-1}(\hat{\rho} - \rho_0)\ \leq \chi^{2}_{p, 1-\alpha} \}$

Modifying from the wald-cloud approach (matrix notation), a random uniform sample can be constructed by:

$$
\hat\rho^{(b)} = \rho_0 + L(\sqrt{c} r^{\frac{1}{p}}Z)
$$

where $c: \chi^2_{p, 1-\alpha}$, $r$ represent a random radius sample from $Unif(0,1)$, $Z$ represents the random unit direction, sample from $N(0,I_p)$ and rescale to have length (norm) 1 (i.e. $||Z|| = 1$)

My intuition on this:

$c$ sets the $1-\alpha$ chi-square cut-off that determines the confidence ellipsoid boundary; $r$ distributes a random radius (between (0,1)) uniformly in all $p$-dimension within the ellipsoid; $Z$ is the standard way to generate uniform random directions on a unit sphere. Geometrically, multiplying these three objects to $L$is like rotating and scaling the ellipsoid.

# Simulating Coverage

Definition of confidence region (CR):

$$
Pr_\rho(\rho \in CR_{1-\alpha}(\hat\rho)) = 1-\alpha 
$$

Frequentist interpretation of confidence region: Over repeated samples, the confidence region (which is random, depends on sample) contains the fixed true $\rho$ with probability $1-\alpha$.

The goal of simulation: check whether nominal and actual coverage of CR are the same, i.e., when $\alpha = 0.05$, do we get actual coverage of 95%?

> **Algorithm: Monte Carlo estimate of coverage of** $CR_{1-\alpha}(\hat{\rho})$
>
> **Input:** $\rho_0$, sample size $n$, significance level $\alpha$, number of replicates $B$
>
> **Output:** Coverage level $CR_{1-\alpha}(\hat\rho)$
>
> $H \leftarrow 0$
>
> **for** $b$ $\leftarrow$ 1 to $B$:
>
> 1\. Sample $\hat\rho^{(b)}$ from $CR_{1-\alpha}(\hat\rho)$ \[Algorithm X\]
>
> 2\. Compute covariance $\hat\Psi(\hat\rho^{(b)})$
>
> 3\. Compute mahalanobis distance $T_b \leftarrow n(\rho_0-\hat{\rho}^{(b)})^{\top}[\hat\Psi(\hat\rho^{(b)})]^{-1}(\rho_0-\hat{\rho}^{(b)})$
>
> 4\. if $T_b \leq \chi^2_{p,1-\alpha}$, then $H \leftarrow H+1$ ;
>
> **return** $H /B$ (actual coverage)

This simulation repeats over different $\rho_0$'s, which are uniformly sampled from the rlkj distribution by under different $\eta$. $n$ is set to $1e6$ to mirror aysmptotic-ness, $\alpha = 0.05, 0.10$, $B=  200$.

Simulation output here

Visualization of
